### 2021-11-19 draft.tex
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@ef6926217c28fafceae6299e600a081ed616414b 🚀

\begin{longtable}[]{@{}l@{}}
\toprule
\endhead
\begin{minipage}[t]{0.05\columnwidth}\raggedright
\textbf{P1}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.05\columnwidth}\raggedright
Within the last several hundred years, human activity have rapidly
changed Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
surface and oceans (resulting in acidification), and both agricultural
and urban development has rapidly reshaped Earth's surface. These the
bulk of this change has occurred within the last several hundred years,
a geological instant, potentially inducing shocks to ecosystems that
could threated their integrity (\textbf{Scheffer?}). As a result
understanding and predicting how ecosystems will change in the future,
\emph{ecological forecasting}, and making making descisions based on
these predictions mitigating the consequences of this change, on
ecosystems has emerged as an imperative for ecology and environmental
science {[}Dietze (2017);{]}.\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

@@

Much of the history of quantitatively modeling ecosystems have been done
in the language of dynamical systems, describing how the value of an
observable state of the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time. It turns
out to be much more effective to, rather than attempt to directly model
\(\vec{x}(t)\) itself, to instead describe how \(\vec{x}\) changes from
one timestep to the next, yielding models in the form of differential
equations in continuous-time settings--\(\frac{dx}{dt} = f(x)\)-- or
difference equations in discrete-time
settings---\(x_t = f(x_{t-1})\)---where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Chen \emph{et al.} 2019).

\begin{longtable}[]{@{}l@{}}
\toprule
\endhead
\begin{minipage}[t]{0.05\columnwidth}\raggedright
\textbf{P3}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.05\columnwidth}\raggedright
However, we run into many problems when aiming to apply this type of
model to empirical data in ecology.\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.05\columnwidth}\raggedright
The initial success of ODE models can be traced back to the larger
program of ontological reductionism, which became the de facto apporoach
model physical sciences after its early success in physics, which, and
by the time ecology was becoming a quantitative science (sometime in the
20th century, depending on who you ask), became the foundation for early
quantitative models in ecology.\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.05\columnwidth}\raggedright
But ecosystems are perhaps the quintessential example of system that
cannot be understood simply by iterative reduction of its components.
Emergent phenomena, mechanisms at different scales, etc.\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.05\columnwidth}\raggedright
Some have been explored in the ecological literature: (1) Some
applications of dynamic models in ecology assume long-run equilibrium.
(2) Stochasticity\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.05\columnwidth}\raggedright
(3) Ecological processes vary across more variables than the tools of
analytic models are suited for. As the number of variables in an
analytic model increases, so does the ability of the scientist to decern
clear relationships between them, and so does overfitting potential.
Curse of dimensionality--- Until the 20th century, no theory of the
gravitational dynamics of more than 2 bodies. Understanding the
gravitational dynamics of more than two planets with any reliability
proved difficult. Using the same models (diffeqs), how could we
adequately predict ecosystems?\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\emph{P4}

The term \emph{ecological forecasting} implicitly creates an analogy
between predicting how ecosystems will change in the future by using the
term ``forecasting''---the most immediate analog is the success story of
forecasting is numerical weather prediction (NWP;
\textbf{Bauer2015QuiRev?}).

Much like ecology, NWP is faced with high-dimensional systems that are
governed by different mechanisms at different scales.

Much as one would not aim to forecast the weather in Quebec by applying
Navier-Stokes. NWP has worked because it incorporates information about
data and meteorological processes collected at difference scales into
models that. Use of computational methods in NWP.

@@

Indeed Marx's most well known quote that ``philosophers have hitherto
only interpreted the world in various ways; the point is to change it.''

and a necessary step toward establishing a just and sustainable world.

Transition to specifics of this thesis.

\begin{figure}
\centering
\includegraphics{./figures/thesisconcept.png}
\caption{thesis concept}
\end{figure}

@@

>>>>

\textbf{P1}

Within the last several hundred years, human activity has rapidly
changed Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrestial surface and its oceans, and both agricultural and urban
development has rapidly reshaped the cover of Earth's surface. These the
bulk of this change has occurred within the last several hundred years,
a geological instant, potentially inducing shocks to ecosystems that
could threated their integrity (\textbf{Scheffer?}). As a result
understanding and predicting how ecosystems will change in the future,
\emph{ecological forecasting}, and making making descisions based on
these predictions mitigating the consequences of this change, on
ecosystems has emerged as an imperative for ecology and environmental
science {[}Dietze (2017);{]}.

@@

@@

Much of the history of quantitatively modeling ecosystems have been done
in the language of dynamical systems, describing how the value of an
observable state of the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time. It turns
out to be much more effective to, rather than attempt to directly model
\(\vec{x}(t)\) itself, to instead describe how \(\vec{x}\) changes from
one timestep to the next, yielding models in the form of differential
equations in continuous-time settings--\(\frac{dx}{dt} = f(x)\)-- or
difference equations in discrete-time
settings--\(x_t = f(x_{t-1})\)--where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Chen \emph{et al.} 2019).

\textbf{P3}

However, we run into many problems when aiming to apply this type of
model to empirical data in ecology.

The initial success of ODE models can be traced back to the larger
program of ontological reductionism, which became the de facto apporoach
model physical sciences after its early success in physics, which, and
by the time ecology was becoming a quantitative science (sometime in the
20th century, depending on who you ask), became the foundation for early
quantitative models in ecology.

But ecosystems are perhaps the quintessential example of system that
cannot be understood simply by iterative reduction of its components.
Emergent phenomena, mechanisms at different scales, etc.

Some have been explored in the ecological literature: (1) Some
applications of dynamic models in ecology assume long-run equilibrium.
(2) Stochasticity

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{2}
\tightlist
\item
  Ecological processes vary across more variables than the tools of
  analytic models are suited for. As the number of variables in an
  analytic model increases, so does the ability of the scientist to
  decern clear relationships between them, and so does overfitting
  potential. Curse of dimensionality--- Until the 20th century, no
  theory of the gravitational dynamics of more than 2 bodies.
  Understanding the gravitational dynamics of more than two planets with
  any reliability proved difficult. Using the same models (diffeqs), how
  could we adequately predict ecosystems?
\end{enumerate}

\textbf{P4}

The term \emph{ecological forecasting} implicitly creates an analogy
between predicting how ecosystems will change in the future by using the
term ``forecasting''---the most immediate analog being the success story
of weather forecasting via numerical weather prediction (NWP).

Although it is become almost hack to complain about the dang weather
forecast being wrong, over the least 50 years the
(\textbf{Bauer2015QuiRev?}).

The success of NWP, and the Earth observations that support it should
serve as a template for development of a system for monitoring Earth's
biodiversity. Much like ecology, NWP is faced with high-dimensional
systems that are governed by different mechanisms at different scales.

NWP has worked because it incorporates information about data and
meteorological processes collected at difference scales into models
that. Use of computational methods in NWP.

Much as one would not aim to forecast the weather in Quebec by applying
Navier-Stokes, forecasting ecological systems must

@@

Indeed Marx's most well known quote that ``philosophers have hitherto
only interpreted the world in various ways; the point is to change it.''
and a necessary step toward establishing a just and sustainable world.

\begin{figure}
\hypertarget{fig:thesis}{%
\centering
\includegraphics{./figures/thesisconcept.png}
\caption{thesis concept}\label{fig:thesis}
}
\end{figure}

\textbf{P6 -- final intro para}

Three major components here: 1) Ecosystem monitoring, 2) Forecasting
using the products of that monitoring, and 3) Choosing the best possible
mitigation strategy.

This flow is outlined in the left panel of fig.~\ref{fig:thesis}

@@

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}
